---
title: "LLMR in 5 minutes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLMR in 5 minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = identical(tolower(Sys.getenv("LLMR_RUN_VIGNETTES", "false")), "true")
)
```

LLMR gives R one interface to many language-model providers. You pick the
provider and model once with `llm_config()`; every other function behaves the
same regardless of which model is behind it.

## 1. Install

```r
install.packages("LLMR")                   # CRAN
# remotes::install_github("asanaei/LLMR")  # development version
```

## 2. Set an API key

You can hand `llm_config()` your key directly as a string, but the safer habit
is to keep it out of your code and let LLMR read it from an environment
variable. For each provider LLMR knows a default variable to look in: it tries
`<PROVIDER>_API_KEY` first, then `<PROVIDER>_KEY` (upper-cased), so Groq reads
`GROQ_API_KEY`, OpenAI reads `OPENAI_API_KEY`, and so on. If you set that
variable, you never pass a key in code at all.

Put the key in your `~/.Renviron` file, one per line:

```
GROQ_API_KEY=...
```

The easiest way to open that file is:

```r
usethis::edit_r_environ()
```

Save it and **restart R**. You can check that R sees the key without printing it:

```r
nzchar(Sys.getenv("GROQ_API_KEY"))   # TRUE once it is set
```

If this is `FALSE`, R cannot see the key yet: check the spelling and that you
restarted the session. A missing key shows up as an authentication error on your
first call, not before.

## 3. Your first call

`llm_config()` selects the model; `call_llm()` sends one message and returns a
response object that prints the text plus a short status line. We use Groq's
open-weight `gpt-oss-20b` here because it is cheap and available to everyone.

```{r}
library(LLMR)

cfg <- llm_config("groq", "openai/gpt-oss-20b", temperature = 0.2)

r <- call_llm(cfg, c(system = "Be concise.", user = "Capital of Mongolia?"))
r                 # prints the text and a [model | finish | tokens | t] line
as.character(r)   # just the text
tokens(r)         # token counts as a list
```

A message is a named character vector; the names are roles (`system`, `user`,
`assistant`). A bare string is treated as a single user turn.

## 4. Apply a model to a data frame

`llm_mutate()` adds model-generated columns to a tibble. The shorthand puts the
new column name and a `glue` prompt template in one argument; `{column}` is
filled from each row.

```{r}
library(tibble)

reviews <- tibble(text = c("The food was cold.",
                           "Absolutely loved it!",
                           "It was fine, nothing special."))

reviews |>
  llm_mutate(
    sentiment = "Reply with one word (positive/negative/neutral): {text}",
    .config   = cfg
  )
```

Alongside the `sentiment` column you also get diagnostic columns
(`sentiment_ok`, `sentiment_finish`, `sentiment_sent`, `sentiment_rec`, ...) so
you can see what succeeded and how many tokens each row used.

## 5. Generative calls over a vector

`llm_fn()` is the lighter-weight sibling of `llm_mutate()`: give it a vector and
a `glue` prompt where `{x}` is each element, and it returns a character vector.

```{r}
countries <- c("Mongolia", "Bolivia", "Chad")

llm_fn(countries,
       prompt  = "Capital city of {x}. Reply with only the city name.",
       .config = cfg)
```

Switching to a different provider or model is a one-line change to
`llm_config()`; nothing else in your code changes.

## 6. Tagged fields and row batching

When you want several fields per row, ask the model to wrap each in a named tag
and pass `.tags`; LLMR parses them into columns. Add `.batch_size` to pack
multiple rows into one request (sent as numbered `<row_i>` blocks and split back
apart), which cuts the number of calls and the repeated instruction overhead.

```{r}
films <- tibble(title = c("Blade Runner", "Amelie", "Parasite", "Spirited Away"))

films |>
  llm_mutate(
    info       = "For the film {title}, give its director and release year.",
    .config    = cfg,
    .tags      = c("director", "year"),
    .batch_size = 2
  )
```

The four films were resolved in two calls (`info_bn = 2`). The `info_batch`,
`info_bn`, and `info_bi` columns record which call each row landed in and its
position within it; the rows always come back in their original order. Prefer
modest batch sizes and `temperature = 0`: batching only pays off when the model
reliably follows the wrapping protocol.

## 7. Embeddings

Embeddings turn text into numeric vectors you can compare. They use a different
kind of model, so you make a config with `embedding = TRUE`; here we use Voyage,
which specializes in embeddings (set `VOYAGE_API_KEY`). `get_batched_embeddings()`
takes a character vector and returns a matrix with one row per text.

```{r}
emb_cfg <- llm_config("voyage", "voyage-3.5-lite", embedding = TRUE)

texts <- c("I love this restaurant.",
           "The food was delicious.",
           "My car broke down today.")

m <- get_batched_embeddings(texts, emb_cfg)
dim(m)   # 3 texts x embedding dimension
```

Closeness in this space tracks meaning. Cosine similarity is high for the two
sentences about food and low for the unrelated one:

```{r}
cosine <- function(a, b) sum(a * b) / sqrt(sum(a * a) * sum(b * b))

cosine(m[1, ], m[2, ])   # food vs food: high
cosine(m[1, ], m[3, ])   # food vs car:  low
```

## 8. Look before you spend, summarize after

`llm_preview()` shows exactly what would be sent, with no API call, so you can
catch a templating or role mistake before paying for it:

```{r}
llm_preview(reviews,
            prompt  = "Reply with one word: {text}",
            .config = cfg)
```

After a run, `llm_usage()` summarizes token totals and outcomes, and
`llm_failures()` lists any rows that failed or were truncated:

```{r}
out <- reviews |>
  llm_mutate(sentiment = "One word for: {text}", .config = cfg)

llm_usage(out)
llm_failures(out)
```

## Where to go next

- **Tidy pipelines and structured output** -- `llm_fn()`, `.tags`, JSON schemas,
  and row batching.
- **Schema-validated output** -- enforcing a JSON shape across providers.
- **Presidential speech analysis** -- a fuller embeddings example with
  clustering.
- **Small experiment** -- factorial designs and parallel execution with
  `call_llm_par()`.
- **Chat sessions** -- stateful multi-turn conversations.
