---
title: "Frictionless Science: The Trolley Dilemma"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frictionless Science: The Trolley Dilemma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = identical(tolower(Sys.getenv("LLMR_RUN_VIGNETTES", "false")), "true")
)
```

Contemporary behavioral research increasingly leverages large language models (LLMs) to simulate human judgments. This vignette illustrates a streamlined methodology for conducting a classical moral philosophy experiment---the Trolley Dilemma---using the `LLMR` package. We bypass the elementary chat functionalities and proceed directly to a vectorised experimental design using `llm_mutate()`. 

For this demonstration, we utilize an open-weights model provided via the Groq API.

```{r setup, message=FALSE, warning=FALSE}
library(LLMR)
library(dplyr)

# Configure an open model endpoint
cfg <- llm_config(
  provider = "groq",
  model    = "llama-3.1-8b-instant"
)
```

## Designing the Experiment

We construct a fundamental stimulus set representing two standard variants of the Trolley Dilemma.

```{r stimuli}
dilemmas <- tibble::tibble(
  condition = c("Switch", "Footbridge"),
  scenario = c(
    "A runaway trolley is heading down the tracks toward five workers who will be killed. You are standing next to a switch. If you pull the switch, the trolley will be diverted onto a side track where it will kill one worker. Do you pull the switch?",
    "A runaway trolley is heading toward five workers. You are standing on a footbridge above the tracks next to a large stranger. If you push the stranger onto the tracks below, his mass will stop the trolley, saving the five workers but killing the stranger. Do you push the stranger?"
  )
)
```

## Vectorised Execution with Soft Structuring

To systematically extract the model's decisions, we deploy `llm_mutate()`. Rather than imposing a rigid JSON schema---which can induce failure modes in some inference endpoints---we instruct the model to use simple XML-like tags. This "soft structuring" approach remains robust across varying provider capabilities.

```{r mutate}
experiment_results <- dilemmas |>
  llm_mutate(
    response = c(
      system = "You are a participant in a moral psychology experiment. Read the scenario and provide a definitive YES or NO decision, followed by a brief rationale. Enclose your decision in <decision>...</decision> tags and your reasoning in <rationale>...</rationale> tags.",
      user = "{scenario}"
    ),
    .config = cfg,
    .tags = c("decision", "rationale")
  )
```

By specifying the `.tags` argument, `LLMR` automatically parses the response string and appends the extracted content as distinct columns in the original dataset.

```{r inspect}
experiment_results |>
  select(condition, decision, rationale) |>
  print(n = Inf)
```

## Conclusion

This approach encapsulates the core objective of `LLMR`: frictionless science. Researchers specify the conditions, articulate a robust prompt, and retrieve a structured dataset ready for statistical evaluation, eliminating intermediate parsing logic and cumbersome loop constructs.
